---
title: "MTG_sep2026_models"
author: "anonymous"
date: "2026-09-23"
output: html_document
---

```{r Setting up}


library (tidyverse)

``` 


Read data and create variables 

```{r data}
data <- read_csv("mtg_sep26.csv", skip_empty_rows = TRUE)

## create gap columns by subtracting from quest_tend

data <- data %>% 
  mutate(
    combined_gap    = combined    - quest_tend,
    viable_gap      = viable      - quest_tend,
    verbal_gap      = verbal      - quest_tend,
    vocal_gap       = vocal_start - quest_tend,
    vegetative_gap  = vegetative  - quest_tend,
    visual_gap      = visual      - quest_tend
  )


## for ease of script re-use:
 timedata <- data


```
# Bayesian inferential modelling - VISUAL cues

```{r GroupStageBayesian visual}

### BAYESIAN MODEL for visual ---------

## setup --------


# package for Bayesian regression modeling
library(brms)

# package for visualization
library(tidybayes)

# package to visualize 
library(bayesplot)

# make Stan run faster
options(mc.cores = parallel::detectCores())

# use cmdstanr (with caution!)
#options(brms.backend = "cmdstanr")

# to change back;
# options(brms.backend = "rstan")






## Full model  --------------------------------



## set weakly informative prior 

priors_model1 <- c(
  set_prior("normal (0,1000)", class = "b"),
  set_prior("normal (0,6000)", class = "Intercept")
)





## Model1 - NOTE ex-gaussian does NOT work with this data set (not easily at all anyway)



model_visual = brm(
  visual_gap ~ proficiency*qType + (1+qType|file) + (1|quest_label),
  prior = priors_model1,
  data = data,
  iter = 4000, warmup = 2000,
  family = gaussian(),
  control = list(adapt_delta = 0.99),
  seed = 55,
  file = "interaction_visual"
)


## posterior predictive check

pp_check (model_visual, nsamples = 100)
pp_check (model_visual, nsamples = 100, type = "boxplot")




##### more pp checks

# Extract 100 posterior predictive draws for all observations
yrep <- posterior_predict(model_visual, nsamples = 100)

# Calculate mean and sd of observed data
mean_obs <- mean(timedata$visual_gap, na.rm=TRUE)
sd_obs <- sd(timedata$visual_gap, na.rm=TRUE)

# Calculate mean and sd for each posterior predictive draw
mean_pred <- apply(yrep, 1, mean)
sd_pred <- apply(yrep, 1, sd)

# Summarize means and SDs of predictions
summary_means <- summary(mean_pred)
summary_sds <- summary(sd_pred)

``` 
```{r visual_summaries}
# Print summaries
print("Observed Data:")
print(paste("Mean:", round(mean_obs, 2), "SD:", round(sd_obs, 2)))

print("Posterior Predictive Means:")
print(summary_means)

print("Posterior Predictive SDs:")
print(summary_sds)

# Optional: check skewness difference 
if(!require(moments)) install.packages("moments")
library(moments)

skew_obs <- skewness(timedata$visual_gap, na.rm=TRUE)
skew_pred <- apply(yrep, 1, skewness)
summary_skew_pred <- summary(skew_pred)

print(paste("Observed skewness:", round(skew_obs, 3)))
print("Posterior Predictive Skewness:")
print(summary_skew_pred)


###------

summary (model_visual)

```
Family: gaussian 
  Links: mu = identity 
Formula: visual_gap ~ proficiency * qType + (1 + qType | file) + (1 | quest_label) 
   Data: data (Number of observations: 65) 
  Draws: 4 chains, each with iter = 4000; warmup = 2000; thin = 1;
         total post-warmup draws = 8000

Multilevel Hyperparameters:
~file (Number of levels: 24) 
                          Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
sd(Intercept)               238.18    124.18    16.26   490.90 1.00     1736     1951
sd(qTypePolar)              167.69    128.07     6.76   469.71 1.00     2999     4011
cor(Intercept,qTypePolar)    -0.20      0.56    -0.97     0.91 1.00     5166     5477

~quest_label (Number of levels: 4) 
              Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
sd(Intercept)   206.50    178.03     8.69   665.89 1.00     2740     3151

Regression Coefficients:
                                   Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
Intercept                            386.16    243.70  -108.47   859.84 1.00     4297     4643
proficiencyIntermediate             -605.84    216.56 -1029.23  -169.57 1.00     4879     4802
qTypePolar                          -121.14    312.83  -729.15   532.61 1.00     4230     4213
proficiencyIntermediate:qTypePolar   454.22    261.79   -63.47   970.88 1.00     5556     6034

Further Distributional Parameters:
      Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
sigma   510.02     57.37   406.01   632.17 1.00     3265     5229

Draws were sampled using sampling(NUTS). For each parameter, Bulk_ESS
and Tail_ESS are effective sample size measures, and Rhat is the potential
scale reduction factor on split chains (at convergence, Rhat = 1).

```{r visual_posteriors_1}
## Posteriors across tasks

posterior_samples <- posterior_samples(model_visual)


# Calculate the group effect for each task
group_effect_open <- posterior_samples$b_proficiencyIntermediate
group_effect_polar <- posterior_samples$b_proficiencyIntermediate + posterior_samples$`b_proficiencyIntermediate:qTypePolar`

# Step 2: Calculate the overall group difference across all tasks by averaging
overall_group_diff_across_tasks <- (group_effect_open + group_effect_polar) / 2

# Step 3: Define custom summary function 
summary_stats <- function(samples) {
  mean_val <- mean(samples)
  lower <- quantile(samples, 0.025)
  upper <- quantile(samples, 0.975)
  prob_above_zero <- mean(samples > 0)
  data.frame(mean = mean_val, lower_95CrI = lower, upper_95CrI = upper, posterior_prob = prob_above_zero)
}

# Step 4: Calculate summary statistics for the overall group difference
results <- summary_stats(overall_group_diff_across_tasks)

results$posterior_prob_flipped <- 1 - results$posterior_prob

print (results)

```
## RESULTS:


    mean lower_95CrI upper_95CrI posterior_prob posterior_prob_flipped
2.5% -376.9669   -703.8009    -40.4225       0.014125               0.985875



```{r visual_posteriors_2}

### Posteriors by task



# Calculate summary statistics for each task
results <- list(
  Polar = summary_stats(group_effect_polar),
  Open = summary_stats(group_effect_open)
)



# View the results
print(results)
# 

``` 
$Polar
          mean lower_95CrI upper_95CrI posterior_prob
2.5% -152.1577   -559.1034    257.5942       0.232125

$Open
          mean lower_95CrI upper_95CrI posterior_prob
2.5% -601.7761   -1018.865   -172.3222         0.0035


```{r visual_posteriors_3}
## QType comparison (open vs polar)

# qType effects by proficiency
qt_beginner <- posterior_samples$b_qTypePolar
qt_intermediate <- posterior_samples$b_qTypePolar + posterior_samples$`b_proficiencyIntermediate:qTypePolar`


results <- list(
  Beginner = summary_stats(qt_beginner),
  Intermediate = summary_stats(qt_intermediate))

print (results)

``` 

$Beginner
          mean lower_95CrI upper_95CrI posterior_prob
2.5% -116.5429   -715.9443    521.9693        0.32325

$Intermediate
         mean lower_95CrI upper_95CrI posterior_prob
2.5% 333.0755   -241.5314     977.373       0.898375


```{r visual_posteriors_4}
# question effect across groups (averaged)


overall_question <- (qt_beginner + qt_intermediate) / 2


results <- summary_stats (overall_question)

results$posterior_prob_flipped <- 1 - results$posterior_prob

print (results)
```
  mean lower_95CrI upper_95CrI posterior_prob posterior_prob_flipped
2.5% 108.2663    -415.567    702.7833        0.68075                0.31925

```{r visual_posteriors_5}

# Interaction effect proficiency*QType

interaction_effect <- posterior_samples$`b_proficiencyIntermediate:qTypePolar`

# Summary
results <- summary_stats(interaction_effect)

results$posterior_prob_flipped <- 1 - results$posterior_prob

print(results)

```
     mean lower_95CrI upper_95CrI posterior_prob posterior_prob_flipped
2.5% 449.6184   -55.20549    971.6661        0.96125                0.03875




# Bayesian inferential modelling - COMBINED (verbal + viable) cues


```{r Bayesian combined}

### BAYESIAN MODEL for combined (verbal and viable) ----------

## setup --------


# package for Bayesian regression modeling
library(brms)

# package for visualization
library(tidybayes)

# package to visualize 
library(bayesplot)

# make Stan run faster
options(mc.cores = parallel::detectCores())

# use cmdstanr (with caution!)
#options(brms.backend = "cmdstanr")

# to change back;
# options(brms.backend = "rstan")





## Gaussian instead of exGaussian due to limited data



## Full model  --------------------------------



## set weakly informative prior 

priors_model1 <- c(
  set_prior("normal (0,1000)", class = "b"),
  set_prior("normal (0,6000)", class = "Intercept")
)





## Model1



model1 = brm(
  combined_gap ~ proficiency*qType + (1+qType|file) + (1|quest_label),
  prior = priors_model1,
  data = data,
  iter = 4000, warmup = 2000,
  family = gaussian(),
  control = list(adapt_delta = 0.99),
  seed = 55,
  file = "m1_full_interaction"
)


## posterior predictive check

pp_check (model1, nsamples = 100)
pp_check (model1, nsamples = 100, type = "boxplot")




##### more pp checks

# Extract 100 posterior predictive draws for all observations
yrep <- posterior_predict(model1, nsamples = 100)

# Calculate mean and sd of observed data
mean_obs <- mean(timedata$combined_gap, na.rm=TRUE)
sd_obs <- sd(timedata$combined_gap, na.rm=TRUE)

# Calculate mean and sd for each posterior predictive draw
mean_pred <- apply(yrep, 1, mean)
sd_pred <- apply(yrep, 1, sd)

# Summarize means and SDs of predictions
summary_means <- summary(mean_pred)
summary_sds <- summary(sd_pred)

# Print summaries
print("Observed Data:")
print(paste("Mean:", round(mean_obs, 2), "SD:", round(sd_obs, 2)))

print("Posterior Predictive Means:")
print(summary_means)

print("Posterior Predictive SDs:")
print(summary_sds)

# Optional: check skewness difference 
if(!require(moments)) install.packages("moments")
library(moments)

skew_obs <- skewness(timedata$combined_gap, na.rm = TRUE)
skew_pred <- apply(yrep, 1, skewness)
summary_skew_pred <- summary(skew_pred)

print(paste("Observed skewness:", round(skew_obs, 3)))
print("Posterior Predictive Skewness:")
print(summary_skew_pred)


###------

summary (model1)
# # 
```

Family: gaussian 
  Links: mu = identity 
Formula: combined_gap ~ proficiency * qType + (1 + qType | file) + (1 | quest_label) 
   Data: data (Number of observations: 96) 
  Draws: 4 chains, each with iter = 4000; warmup = 2000; thin = 1;
         total post-warmup draws = 8000

Multilevel Hyperparameters:
~file (Number of levels: 24) 
                          Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
sd(Intercept)              1387.86    237.84   977.71  1912.19 1.00     2214     4051
sd(qTypePolar)             1232.16    260.90   764.29  1791.58 1.00     2880     4335
cor(Intercept,qTypePolar)    -0.96      0.04    -1.00    -0.84 1.00     4167     3896

~quest_label (Number of levels: 4) 
              Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
sd(Intercept)   760.60    489.75   187.91  2002.46 1.00     2572     3700

Regression Coefficients:
                                   Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
Intercept                           2397.96    644.28  1001.63  3564.33 1.00     3161     3797
proficiencyIntermediate             -283.13    483.56 -1208.79   689.09 1.00     2566     4194
qTypePolar                          -996.34    707.74 -2250.15   582.98 1.00     3849     5136
proficiencyIntermediate:qTypePolar   -53.41    486.99 -1020.85   880.20 1.00     3349     4821

Further Distributional Parameters:
      Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
sigma   880.68     82.21   734.88  1056.85 1.00     4876     5473

Draws were sampled using sampling(NUTS). For each parameter, Bulk_ESS
and Tail_ESS are effective sample size measures, and Rhat is the potential
scale reduction factor on split chains (at convergence, Rhat = 1).

summary (model1)





```{r combined_posterior1}


posterior_samples <- posterior_samples(model1)


# Step 1: Calculate the group effect for each task
group_effect_open <- posterior_samples$b_proficiencyIntermediate
group_effect_polar <- posterior_samples$b_proficiencyIntermediate + posterior_samples$`b_proficiencyIntermediate:qTypePolar`

# Step 2: Calculate the overall group difference across all tasks by averaging
overall_group_diff_across_tasks <- (group_effect_open + group_effect_polar) / 2

# Step 3: Define custom summary function 
summary_stats <- function(samples) {
  mean_val <- mean(samples)
  lower <- quantile(samples, 0.025)
  upper <- quantile(samples, 0.975)
  prob_above_zero <- mean(samples > 0)
  prob_below_zero <- 1 - prob_above_zero
  evidence_ratio1 <- prob_above_zero / prob_below_zero
  evidence_ratio2 <- prob_below_zero / prob_above_zero
  data.frame(mean = mean_val, lower_95CrI = lower, upper_95CrI = upper, posterior_prob = prob_above_zero, ER1 = evidence_ratio1, ER2 = evidence_ratio2)
}

# Step 4: Calculate summary statistics for the overall group difference
results <- summary_stats(overall_group_diff_across_tasks)

results$posterior_prob_flipped <- 1 - results$posterior_prob

print (results)

```

## RESULTS: Overall effect of proficiency
# 
mean lower_95CrI upper_95CrI posterior_prob       ER1      ER2 posterior_prob_flipped
2.5% -309.8345   -938.0307    328.8906       0.167125 0.2006604 4.983545               0.832875

```{r combined_posterior2}

## group effect open questions only

results <- summary_stats(group_effect_open)

results$posterior_prob_flipped <- 1 - results$posterior_prob

print (results)

```



mean lower_95CrI upper_95CrI posterior_prob       ER1      ER2 posterior_prob_flipped
2.5% -283.128   -1208.792    689.0944       0.273125 0.3757524 2.661327               0.726875

```{r combined_posterior3}

## group effect polar questions only

results <- summary_stats(group_effect_polar)

results$posterior_prob_flipped <- 1 - results$posterior_prob

print (results)

```

mean lower_95CrI upper_95CrI posterior_prob      ER1    ER2 posterior_prob_flipped
2.5% -336.5409    -933.938    265.5146          0.128 0.146789 6.8125                  0.872


```{r combined_posterior4}


### Posteriors for group effect by question type (open vs polar)


# Calculate summary statistics for each task
results <- list(
  Polar = summary_stats(group_effect_polar),
  Open = summary_stats(group_effect_open)
)



# View the results
print(results)

```
$Polar
          mean lower_95CrI upper_95CrI posterior_prob      ER1    ER2
2.5% -336.5409    -933.938    265.5146          0.128 0.146789 6.8125

$Open
         mean lower_95CrI upper_95CrI posterior_prob       ER1      ER2
2.5% -283.128   -1208.792    689.0944       0.273125 0.3757524 2.661327



```{r combined_posterior5}

## QType comparison by group

# qType effects
qt_beginner <- posterior_samples$b_qTypePolar
qt_intermediate <- posterior_samples$b_qTypePolar + posterior_samples$`b_proficiencyIntermediate:qTypePolar`


results <- list(
  Beginner = summary_stats(qt_beginner),
  Intermediate = summary_stats(qt_intermediate))

print (results)

``` 
$Beginner
          mean lower_95CrI upper_95CrI posterior_prob        ER1      ER2
2.5% -996.3403   -2250.153    582.9806          0.089 0.09769484 10.23596

$Intermediate
          mean lower_95CrI upper_95CrI posterior_prob       ER1      ER2
2.5% -1049.753   -2420.981    597.4666        0.09125 0.1004127 9.958904

```{r combined_posterior6}
# question effect across groups


overall_question <- (qt_beginner + qt_intermediate) / 2


results <- summary_stats (overall_question)

results$posterior_prob_flipped <- 1 - results$posterior_prob

print (results)
```

         mean lower_95CrI upper_95CrI posterior_prob        ER1      ER2 posterior_prob_flipped
2.5% -1023.047   -2224.543    543.2576        0.08175 0.08902804 11.23242                0.91825

```{r combined_posterior7}
# interaction term proficiency*qytpe

# Interaction effect posterior samples
interaction_effect <- posterior_samples$`b_proficiencyIntermediate:qTypePolar`

# Summary
results <- summary_stats(interaction_effect)

results$posterior_prob_flipped <- 1 - results$posterior_prob

print(results)
```


mean lower_95CrI upper_95CrI posterior_prob      ER1      ER2 posterior_prob_flipped
2.5% -53.41296    -1020.85     880.196       0.459125 0.848856 1.178056               0.540875
